---
title: "LCWMD Chloride Data Analysis"
author: "Curtis C. Bohlen"
date: "3/2/2021"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      fig.align = 'center',
                      fig.width = 5, fig.height = 4,
                      collapse = TRUE, comment = "#>")
```


# Load Libraries
(code omitted)
```{r echo = FALSE}
library(fitdistrplus)  # Loads MASS, which has `select()`, so load first
library(tidyverse)
library(readxl)

library(mblm)

#library(lme4)
library(emmeans)

library(CBEPgraphics)
load_cbep_fonts()
theme_set(theme_cbep())
```


# Load Data
```{r warning = FALSE}
chl_data <- read_excel("Chloride Regression Statistics -2020 End Of Year.xlsx", 
                       sheet = 'SpC&Cl Data',
                       col_types = c("text", "text", "text", 
                                     "date", "date", "skip", "numeric", 
                                     "skip", "numeric", "skip", "skip", 
                                     "skip", "skip", "skip"),
                       .name_repair = 'universal') %>%
  rename(Chloride = Result) %>%
  mutate(Month = as.numeric(format(Sample.Date, format = '%m')),
         Month = factor(Month, levels = 1:12, labels = month.abb),
         Year  = factor(format(Sample.Date, format = '%Y'))) %>%
  mutate(Site.ID = fct_reorder(Site.ID, Chloride))
```

# Flag Outliers
```{r}
chl_data <- chl_data %>%
  mutate(Outlier = Chloride > 1500)
```

# Plot Data
```{r fig.width = 7}
ggplot(chl_data, aes(x = SpC, y = Chloride, color = Sample.Type)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = 'lm', se = FALSE) +
  scale_color_manual(values = cbep_colors())
```




# Simple Linear Regression
```{r}
all_lm <- lm(Chloride ~ SpC, data = chl_data)
summary(all_lm)
```


```{r}
plot(all_lm)
```

#Reviewing Residuals
So the regression diagnostics clearly show the residuals deviate substantially
from a normal distribution.  The strongest deviations are from our two outliers
(# 721 and 722), but even in their absence, the residuals are kurtotic and
slightly skewed.  There is also a small scale - location relationship, which
is probably to be expected.

```{r}
df <- tibble(r = resid(all_lm))

ggplot(df, aes(r)) +
  geom_histogram(aes(y = ..density..)) +
  geom_function(fun = dnorm,
                args = list(mean = mean(df$r), 
                sd = sd(df$r)))
```
It suggests a different model form may be more useful

# Cullen and Frey Moments Graph
```{r}
descdist(df$r, boot = 1000)
```

Which is little help.  That suggest the distributional properties are not too
far from a gamma distribution, but the gamma distribution is defined
over the positive real numbers, and our residuals (by definition) have mean 
zero.

# Omitting Outliser
```{r}
wo_lm <- lm(Chloride ~ SpC, data = chl_data, subset = ! Outlier)
summary(wo_lm)
```

```{r}
plot(wo_lm)
```

It's still not a great model, give the structure of the errors.  The kurtotic 
pattern of the residuals remains.  It suggests a model that has left out some
important predictor variables


# Robust Regression
We are using a Modified "Thiel-Sen Estimator" which is highly resistant to
outliers.
```{r}
robust_lm <- mblm(Chloride ~ SpC, data = chl_data)
summary(robust_lm)
```

# Compare Simple Linear Models
```{r}
df <- tibble(SpC = seq(0, 3500, 100))


p_all <- predict(all_lm, newdata = df, interval = "prediction")
p_wo  <- predict(wo_lm, newdata = df, interval = "prediction")
p_robust <- predict(robust_lm, newdata = df, interval = "prediction")

df <- df %>%
  mutate( pp_all = p_all[,1], pp_wo = p_wo[,1], pp_robust = p_robust[,1]) %>%
  mutate( low_all = p_all[,2], low_wo = p_wo[,2], low_robust = p_robust[,2]) %>%
  mutate( hi_all = p_all[,3], hi_wo = p_wo[,3], hi_robust = p_robust[,3])
```


```{r}
ggplot(df, (aes(x = SpC))) +
  geom_point(data = chl_data, 
             mapping = aes(SpC, Chloride), alpha = 0.1) +
  #geom_smooth(data = chl_data, mapping = aes(SpC, Chloride), 
  #            method = lm, formula = y~ poly(x,2), se = FALSE) +
  geom_line(aes(y = pp_all), color = 'black') +
  geom_ribbon(aes(ymin = low_all, ymax = hi_all), alpha = 0.1, fill = 'black') +
  geom_line(data = df, mapping = aes(SpC, pp_wo), color = 'red') +
  geom_ribbon(aes(ymin = low_wo, ymax = hi_wo), alpha = 0.1, fill = 'red') +
  geom_line(data = df, mapping = aes(SpC, pp_robust), color = 'blue') +
  geom_ribbon(aes(ymin = low_robust, ymax = hi_robust), alpha = 0.1, fill = 'blue') +
  
  theme(legend.position = 'none') +
  xlab(expression('Specific Conductance (' * mu * S/cm *')')) +
  ylab('Chloride (mg/l)')
```
Note that the robust regression (blue)line is parallel to, but somewhat below the
other two regression lines. Error bands (95% confidence intervals of
prediction) overlap.  Many observations fall outside of those 95% confidence 
intervals,  especially on the high side, suggesting some process may be
generating unusually high chloride values.  The NUMBER of high values is not
extreme, but they are further from the 95% confidence band than we might expect.
(As also  shown bu the histogram of deviations, above).

# Polynomial Regressions
In reviewing preliminary models, I observed that the relationship between
conductivity and chlorides is not quite linear. This actually corresponds to 
theory, which suggests we should see a shallow convex up relationship.

We fit nested polynomial models to test if that's reasonable.

```{r}
poly_lm_1 <- lm(Chloride ~ SpC + I(SpC^2) + I(SpC^3) + I(SpC^4), 
              data = chl_data,
              subset = ! Outlier)

anova(poly_lm_1)
```

```{r}
summary(poly_lm_1)
```
Note that although the polynomial terms are significant by F test, they are not 
individually significant by T test, probably because of co-linearity.  We can 
fit orthogonal polynomials using the `poly()` function.

Although a third order polynomial fits better than a second order polynomial, 
we stick with a second order polynomial for its simplicity. a Very lightly
smoothed GAM might also work.

```{r}
poly_lm <- lm(Chloride ~ poly(SpC,2), 
              data = chl_data,
              subset = ! Outlier)
anova(poly_lm)
summary(poly_lm)
```

That fits better than the linear model
```{r}
anova(wo_lm, poly_lm)

```

Theory suggests concentration of solutes should combine a linear term with a 
term raised to the three half power. 

fitting a square root term, and this model actually fits
far better than the straight polynomial model.
```{r}
theory_lm <-  lm(Chloride ~ SpC + I(SpC^1.5), 
              data = chl_data,
              subset = ! Outlier)

sqrt_lm <- lm(Chloride ~ SpC + sqrt(SpC), 
              data = chl_data,
              subset = ! Outlier)
anova(wo_lm, sqrt_lm, theory_lm, poly_lm)
```

The square root model fits the best. Since all three models have the same 
degrees of freedom, there is no formal statistical test of what "better" means here.  

# Compare Polynomial Models
```{r}
df <- tibble(SpC = seq(0, 3500, 100))

p_wo  <- predict(wo_lm, newdata = df, interval = "prediction")
p_poly  <- predict(poly_lm, newdata = df, interval = "prediction")
p_sqrt <- predict(sqrt_lm, newdata = df, interval = "prediction")

df <- df %>%
  mutate( pp_wo = p_wo[,1], pp_poly = p_poly[,1], pp_sqrt = p_sqrt[,1]) %>%
  mutate( low_wo = p_wo[,2], low_poly = p_poly[,2], low_sqrt = p_sqrt[,2]) %>%
  mutate( hi_wo = p_wo[,3], hi_poly = p_poly[,3], hi_sqrt = p_sqrt[,3])
```


```{r fig.width = 7, fig.height = 5}
ggplot(df, (aes(x = SpC))) +
  geom_point(data = chl_data, 
             mapping = aes(SpC, Chloride), alpha = 0.1) +
  #geom_smooth(data = chl_data, mapping = aes(SpC, Chloride), 
  #            method = lm, formula = y~ poly(x,2), se = FALSE) +
  geom_line(aes(y = pp_wo), color = 'black') +
  geom_ribbon(aes(ymin = low_wo, ymax = hi_wo), alpha = 0.1, fill = 'black') +
  geom_line(data = df, mapping = aes(SpC, pp_poly), color = 'red') +
  geom_ribbon(aes(ymin = low_poly, ymax = hi_poly), alpha = 0.1, fill = 'red') +
  #geom_line(data = df, mapping = aes(SpC, pp_sqrt), color = 'blue') +
  #geom_ribbon(aes(ymin = low_sqrt, ymax = hi_sqrt), alpha = 0.1, fill = 'blue') +
  
  theme(legend.position = 'none') +
  xlab(expression('Specific Conductance (' * mu * S/cm *')')) +
  ylab('Chloride (mg/l)')
```

#  Multiple Regression Models
The wide outliers suggest our models are incomplete, omitting some
predictor that could help explain extreme values.  We can try to look at
that by adding the few predictors we have available.

We chose not to include the Site term in the interactions, as we have few 
observations from some sites in several years, leading to risk of bias.
We might be better off fitting a hierarchical model, with random intercepts and
slopes by Site.  We do not pursue that in this quick and dirty analysis.

## Maximum Model
```{r}
full_lm <- lm(Chloride ~ Year + Site.ID + (poly(SpC,2) + Sample.Type)^2, 
              data = chl_data,
              subset = ! Outlier)
anova(full_lm)
```

## Step Model Selection
All terms are "significant", but it's worth wondering if they are informative.
We test with `step()`, which compares models based on AIC.
```{r}
step_lm <- step(full_lm)
anova(step_lm)


```
The `step()` function retains all terms from the full model 

```{r}
plot(step_lm)
```
So while the "optimal"  model includes site and sample type terms, as well 
a interactions with SpC, the model weaknesses are little improved.  The 
model still does not do a very good job of predicting outliers.

```{r}
summary(step_lm)
```

Note that in this formulation, the second order polynomial term is involved with 
a large and significant interaction. The polynomial may only be needed to model 
storm event samples. This deserves further exploration, by plotting data and
predictions, by Sample Type, but we do not bother.  At a guess, the curvature
in part reflects the fact that both chlorides and conductivity tend to be 
slightly lower during storm events.

## Marginal Means and Trends
```{r fig.width = 7}
emmip(step_lm,  Sample.Type ~  SpC, at = list(SpC = 10:300 * 10))
```
That's some indications that the Conductivity:Chloride relationship differs by
Sample Type. note, however, that we actually have relatively little data for
storm samples from the upper ranges, suggesting the main effect is that the 
conductivity - chloride slope may be less steep for the Storm samples.

```{r}
final_lm <- lm(Chloride ~ SpC + I(SpC^2) +
                 Site.ID + Year + 
                 Sample.Type + SpC:Sample.Type, 
              data = chl_data,
              subset = ! Outlier)
anova(final_lm, step_lm)
```
So the slightly simplified model (omitting interaction terms for the curvature)
is slightly worse than the full model.

```{r}
emmip(final_lm,  Sample.Type ~ SpC, at = list(SpC = 10:300 * 10))
```

WE can regain the high recent salinity conclusions we found for State of the Bay.
```{r}
emms <- summary(emmeans(final_lm, "Year"))

plot(emms)
```

And can examine differences among sites. These have been sorted by median 
chlorides.  I fear there's some bias here because of different sampling
histories that can't be fully resolved.  But note only the two (or maybe
three) highest salinity sites are significantly different from the others.

```{r}
emms <- summary(emmeans(final_lm, "Site.ID"))

plot(emms)
```


